### 2024-04-05 thesis/sections/4-datasets.tex
structure and comments
describe experimental setups
attention

\section{Creation of the dataset}
\label{sec:creation-dataset}
This research investigates how agents use referring expressions to discriminate objects seen in images based on their relations.
For that reason, the original CLEVR framework offers too little control over how a new dataset is created.
Especially, which objects and in which attributes they share with each other in each image can't be controlled.
Following, we extended the framework for generating new datasets.\footnote{\href{https://github.com/DominikKuenkele/MLT\_Master-Thesis\_clevr-dataset-gen}{https://github.com/DominikKuenkele/MLT\_Master-Thesis\_clevr-dataset-gen}}
By this, the objects in the generated images are controlled to have different human-recognizable attributes, namely the \emph{shape}, \emph{size} and \emph{color}.
These attributes also correspond to referring expressions in natural language such as English.
% To simplify the generation and the succeeding training of the models, we only focused on the attributes with a high impact on the appearance of the object, namely the \emph{shape}, \emph{size} and \emph{color}.
The \emph{material} is always the same for all objects in a generated image.
There were three main extensions to the framework:
% SD: Too informal
% SD: We create a new dataset based on the CLEVR framework where we control the appearance of scenes in a referring expression task.
% SD: The scenes are controlled by human-recognisable attributes of object such as object shape, colour and type. These attributes also correspond to referring expressions in natural language such as English.
% SD: We create two contexts of scenes, one with two objects and one with five. 
% SD: We also control for the number of attributes shared between the target and the distractor.
% DK: creation of dataset later (here only the general rules), (added; done)

@@

Two datasets following these rules are created.
The Dale-2 dataset contains one target object and one distractor (see Figure \ref{fig:clevr-dale-2}), while the Dale-5 dataset contains one target object and exactly four distractors.
Consider Figure \ref{fig:clevr-dale-5}, with the target object being the \emph{large purple cylinder}. The large purple sphere shares the size and color, the two cubes only share the size, and the small turquoise sphere doesn't share any attribute.
% SD: It’s actually the reverse in which descriptions are generated, we start left to right: large purple (are shared), the last attribute must thus be unique. Generation proceeds in the opposite order, one would just say sphere. We should clarify this. Now it is very hard to understand.
% DK: That's true for the generation of the descriptions. This is explained in chapter 4.2.1 where the captions are needed for the training as input. As said before, the algorithm was not directly used and is just the reason. (done)

@@

\cmtDK{probabilities of shared attributes for both Dale-2 and Dale-5}

>>>>

\section{Creation of the datasets}
\label{sec:creation-dataset}
This research investigates how agents use referring expressions to discriminate objects seen in images based on their relations.
For that reason, the original CLEVR framework offers too little control over how a new dataset is created.
Especially, which objects and in which attributes they share with each other in each image can't be controlled.
Following, we extended the framework for generating new datasets.\footnote{\href{https://github.com/DominikKuenkele/MLT\_Master-Thesis\_clevr-dataset-gen}{https://github.com/DominikKuenkele/MLT\_Master-Thesis\_clevr-dataset-gen}}
By this, the objects in the generated images are controlled to have different human-recognizable attributes, namely the \emph{shape}, \emph{size} and \emph{color}.
These attributes also correspond to referring expressions in natural language such as English.
% To simplify the generation and the succeeding training of the models, we only focused on the attributes with a high impact on the appearance of the object, namely the \emph{shape}, \emph{size} and \emph{color}.
The \emph{material} is always the same for all objects in a generated image.
There were three main extensions to the framework:
% SD: Too informal
% SD: We create a new dataset based on the CLEVR framework where we control the appearance of scenes in a referring expression task.
% SD: The scenes are controlled by human-recognisable attributes of object such as object shape, colour and type. These attributes also correspond to referring expressions in natural language such as English.
% SD: We create two contexts of scenes, one with two objects and one with five. 
% SD: We also control for the number of attributes shared between the target and the distractor.
% DK: creation of dataset later (here only the general rules), (added; done)

@@

Two datasets following these rules are created.
The 'Dale-2' dataset contains one target object and one distractor (see Figure \ref{fig:clevr-dale-2}), while the Dale-5 dataset contains one target object and exactly four distractors.
Consider Figure \ref{fig:clevr-dale-5}, with the target object being the \emph{large purple cylinder}. The large purple sphere shares the size and color, the two cubes only share the size, and the small turquoise sphere doesn't share any attribute.
% SD: It’s actually the reverse in which descriptions are generated, we start left to right: large purple (are shared), the last attribute must thus be unique. Generation proceeds in the opposite order, one would just say sphere. We should clarify this. Now it is very hard to understand.
% DK: That's true for the generation of the descriptions. This is explained in chapter 4.2.1 where the captions are needed for the training as input. As said before, the algorithm was not directly used and is just the reason. (done)

@@

### 2024-04-05 thesis/sections/5.3-reference-resolution.tex
learning curves, architectures
rename
attention

To do this, the model combines a representation of the image and a representation of the target object.
More specifically, the image is encoded using the \emph{image encoder} submodule, omitting the last \emph{max pool} layer.
By this, the resulting matrix has 128\times14\times14 dimension, corresponding to the 14\time14 regions.
For the target object, the incremental RE is passed through an LSTM as described before.
Both encodings are passed through two linear layers separately to project them to the projection dimension $p$, followed by the $tanh$ function.
Following, the dot product is calculated between the projected image and the projected attributes which results in a 14\times14 matrix.
This emphasizes the correlation between the attributes and each region.
A high dot product for a region indicates a high correlation between the attributes and the specific region, while a low dot product indicates the opposite.
Training a model like this should therefore highlight the regions in the image that are described by the referring expression, namely the region around the target object.
To calculate the loss, the $softmax$ function is applied over the prediction and compared to the ground truth matrix $A$ using the \emph{binary cross entropy}.
Hereby, the models converged faster and were trained for 20 epochs with a learning rate of $2\times10^{-4}$.
An $LSTM_e=15$ and $LSTM_o=1500$ showed the best results.
Projection sizes $p=\{50,100,300,500,1000\}$ are compared.

@@

\begin{figure}[ht]
    \centering
    \subfigure['Dale-2', train split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_dale-2_train.png}
        \label{fig:visualizations_dale-2_train}
    }
    \subfigure['Dale-2', test split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_dale-2_test.png}
        \label{fig:visualizations_dale-2_test}
    }
    \subfigure['Dale-5', train split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_dale-5_train.png}
        \label{fig:visualizations_dale-5_train}
    }
    \subfigure['Dale-5', test split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_dale-5_test.png}
        \label{fig:visualizations_dale-5_test}
    }
    \subfigure['CLEVR color', train split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_colour_train.png}
        \label{fig:visualizations_colour_train}
    }
    \subfigure['CLEVR color', test split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_colour_test.png}
        \label{fig:visualizations_colour_test}
    }
    \caption{Examples of the models' predictions on the 'Dale' and 'CLEVR color' datasets. The green circle represents the target object, the red object the model's prediction. Notice that the images are shown in the original size while the model processes them normalized. Hence, the distances don't correspond to the axes.}
    \label{fig:visualizations}
\end{figure}

An interesting pattern appears when doing a qualitative analysis of the models' predictions.
Here, the predicted coordinates compared to the ground truth coordinates are visualized (Figure \ref{fig:visualizations}), using the best performing model with $e_i=300$, $c=2048$, $LSTM_o=1500$ and $LSTM_e=15$.
Figure \ref{fig:visualizations_dale-2_train} shows random examples of predictions for images in the train dataset of Dale-2.
The green circle shows the ground truth center coordinates of the target object, while the red circle shows the prediction of the model.
As can be seen, the predictions are very precise.
Figure \ref{fig:visualizations_complete} combines the predictions and ground truths across all images in the train dataset.
This shows general patterns of the models predictions over the complete dataset.
Here, all predicted coordinates are placed as red circles into the image, while all ground truth coordinates are placed as green circles.
For the predictions on the 'Dale-2' dataset, the resulting shape is a rhombus, which reflects that all objects are placed usually central into the scene (see Figure \ref{fig:visualizations_dale-2_complete_train}).
As expected the green and red rhombus align mostly in the same area for the train split of the 'Dale-2' dataset.
% SD: But this image does not show clearly whether the circles for the same object match. You could calculate the average error in distance between the ground truth and predicted coordinates.
% DK: that is done in the quantitative analysis (mean square error above). The visualization should show general learned patterns of the model (as in this case predictions toard the center)
Figures \ref{fig:visualizations_dale-5_train} and \ref{fig:visualizations_dale-5_complete_train} show the same patterns for the 'Dale-5' dataset, while Figures \ref{fig:visualizations_colour_train} and \ref{fig:visualizations_colour_complete_train} show it for the 'CLEVR color' dataset.

@@

\begin{figure}[ht]
    \centering
    \subfigure['Dale-2', train split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_dale-2_train_complete.png}
        \label{fig:visualizations_dale-2_complete_train}
    }
    \subfigure['Dale-2', test split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_dale-2_test_complete.png}
        \label{fig:visualizations_dale-2_complete_test}
    }
    \subfigure['Dale-5', train split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_dale-5_train_complete.png}
        \label{fig:visualizations_dale-5_complete_train}
    }
    \subfigure['Dale-5', test split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_dale-5_test_complete.png}
        \label{fig:visualizations_dale-5_complete_test}
    }
    \subfigure['CLEVR color', train split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_colour_train_complete.png}
        \label{fig:visualizations_colour_complete_train}
    }
    \subfigure['CLEVR color', test split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_colour_test_complete.png}
        \label{fig:visualizations_colour_complete_test}
    }
    \caption{The images show all the model's predictions (red dots) and the ground truths (green dots) combined for the 'Dale' and 'CLEVR color' datasets}
    % SD: I Haven't checked the earlier captions, but the captions should be informative in the sense that one does not need to look for the text to understand them. Hence, include a brief summary of what each figure contains.
    % DK: (done)
    \label{fig:visualizations_complete}
\end{figure}

@@

>>>>

To do this, the model combines a representation of the image and a representation of the target object.
More specifically, the image is encoded using the \emph{image encoder} submodule, omitting the last \emph{max pool} layer.
By this, the resulting matrix has 128\times14\times14 dimension, corresponding to the 14\times14 regions.
For the target object, the incremental RE is passed through an LSTM as described before.
Both encodings are passed through two linear layers separately to project them to the projection dimension $p$, followed by the $tanh$ function.
Following, the dot product is calculated between the projected image and the projected attributes which results in a 14\times14 matrix.
This emphasizes the correlation between the attributes and each region.
A high dot product for a region indicates a high correlation between the attributes and the specific region, while a low dot product indicates the opposite.
Training a model like this should therefore highlight the regions in the image that are described by the referring expression, namely the region around the target object.
To calculate the loss, the $softmax$ function is applied over the prediction and compared to the ground truth matrix $A$ using the \emph{binary cross entropy}.
Figure \ref{fig:attention_predictor_architecture} shows the architecture.
Hereby, the models converged faster and were trained for 20 epochs with a learning rate of $2\times10^{-4}$.
An $LSTM_e=15$ and $LSTM_o=1500$ showed the best results.
Projection sizes $p=\{50,100,300,500,1000\}$ are compared.

\begin{figure}[ht]
    \centering
    \includegraphics[width=.25\linewidth]{figures/arch_attention_predictor.png}
    \caption{Simplified architecture of the attention predictor}
    \label{fig:attention_predictor_architecture}
\end{figure}

@@

\begin{figure}[p]
    \centering
    \subfigure['Dale-2', train split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_dale-2_train.png}
        \label{fig:visualizations_dale-2_train}
    }
    \subfigure['Dale-2', test split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_dale-2_test.png}
        \label{fig:visualizations_dale-2_test}
    }
    \subfigure['Dale-5', train split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_dale-5_train.png}
        \label{fig:visualizations_dale-5_train}
    }
    \subfigure['Dale-5', test split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_dale-5_test.png}
        \label{fig:visualizations_dale-5_test}
    }
    \subfigure['CLEVR color', train split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_colour_train.png}
        \label{fig:visualizations_colour_train}
    }
    \subfigure['CLEVR color', test split]{
        \includegraphics[width=.839\linewidth]{figures/visualization_colour_test.png}
        \label{fig:visualizations_colour_test}
    }
    \caption{Examples of the models' predictions on the 'Dale' and 'CLEVR color' datasets. The green circle represents the target object, the red object the model's prediction. Notice that the images are shown in the original size while the model processes them normalized. Hence, the distances don't correspond to the axes.}
    \label{fig:visualizations}
\end{figure}

An interesting pattern appears when doing a qualitative analysis of the models' predictions.
Here, the predicted coordinates compared to the ground truth coordinates are visualized (Figure \ref{fig:visualizations}), using the best performing model with $e_i=300$, $c=2048$, $LSTM_o=1500$ and $LSTM_e=15$.
Figure \ref{fig:visualizations_dale-2_train} shows random examples of predictions for images in the train dataset of 'Dale-2'.
The green circle shows the ground truth center coordinates of the target object, while the red circle shows the prediction of the model.
As can be seen, the predictions are very precise.
Figure \ref{fig:visualizations_complete} combines the predictions and ground truths across all images in the train dataset.
This shows general patterns of the models predictions over the complete dataset.
Here, all predicted coordinates are placed as red circles into the image, while all ground truth coordinates are placed as green circles.
For the predictions on the 'Dale-2' dataset, the resulting shape is a rhombus, which reflects that all objects are placed usually central into the scene (see Figure \ref{fig:visualizations_dale-2_complete_train}).
As expected the green and red rhombus align mostly in the same area for the train split of the 'Dale-2' dataset.
% SD: But this image does not show clearly whether the circles for the same object match. You could calculate the average error in distance between the ground truth and predicted coordinates.
% DK: that is done in the quantitative analysis (mean square error above). The visualization should show general learned patterns of the model (as in this case predictions toard the center)
Figures \ref{fig:visualizations_dale-5_train} and \ref{fig:visualizations_dale-5_complete_train} show the same patterns for the 'Dale-5' dataset, while Figures \ref{fig:visualizations_colour_train} and \ref{fig:visualizations_colour_complete_train} show it for the 'CLEVR color' dataset.

@@

\begin{figure}[ht!]
    \centering
    \subfigure['Dale-2', train split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_dale-2_train_complete.png}
        \label{fig:visualizations_dale-2_complete_train}
    }
    \subfigure['Dale-2', test split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_dale-2_test_complete.png}
        \label{fig:visualizations_dale-2_complete_test}
    }
    \subfigure['Dale-5', train split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_dale-5_train_complete.png}
        \label{fig:visualizations_dale-5_complete_train}
    }
    \subfigure['Dale-5', test split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_dale-5_test_complete.png}
        \label{fig:visualizations_dale-5_complete_test}
    }
    \subfigure['CLEVR color', train split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_colour_train_complete.png}
        \label{fig:visualizations_colour_complete_train}
    }
    \subfigure['CLEVR color', test split]{
        \includegraphics[width=0.42\linewidth]{figures/visualization_colour_test_complete.png}
        \label{fig:visualizations_colour_complete_test}
    }
    \caption{The images show all the model's predictions (red dots) and the ground truths (green dots) combined for the 'Dale' and 'CLEVR color' datasets}
    % SD: I Haven't checked the earlier captions, but the captions should be informative in the sense that one does not need to look for the text to understand them. Hence, include a brief summary of what each figure contains.
    % DK: (done)
    \label{fig:visualizations_complete}
\end{figure}

@@

A central problem in the models' predictions is therefore the focussing on precise locations in the image and connecting them to the referring expression.
The \textbf{attention reference resolver} is meant to solve this problem in two ways.
First, the predictions must not be as precise, as only regions in a 14\times14 grid need to be identified.
Second, the referring expressions are naturally combined with each region in the image.
While the \emph{coordinate reference resolver} only combines the complete feature map with the encoded referring expression, the \emph{attention reference resolver} calculates the similarity of each region with the referring expression separately.
This should connect both encodings easier.
As Table \ref{tab:results:attention-reference-resolver} shows, this hypothesis seems to be confirmed.
Across all datasets, the model is able focus on the correct region in the image with high precision.
On the 'Dale-2' dataset, the model is able to make 96,57\% of its predictions in the correct regions around the target object with a projection size of $p=500$.
The worst result with $p=300$ still only lies around 3\% points below and therefore identifies the target object almost always correctly.
These results still hold, when more distractors and more complex referring expressions are introduced.
On the 'Dale-5' dataset, the model can achieve similar performances, with a probability mass of 93,61\% with $p=300$, while three of the remaining configurations stay above 92\%.
When the \emph{color} is the only discriminating attribute, the model achieves again similar results to the 'Dale-2' dataset, despite the higher number of distractors of up to 10.
The best performance of 95,33\% is achieved with $p=100$, but all configurations beat are almost reach 94\%.
In conclusion, the models are able to easily understand the referring expression and to ground them in the target object in the image.
While more complex referring expressions challenge the model slightly more, it still can focus on the correct object.

\begin{table}[ht]
    \centering
    \begin{tabular}{c|c|c|c}
        \toprule
               & \multicolumn{1}{c}{\textbf{Dale-2}} & \multicolumn{1}{c}{\textbf{Dale-5}} & \multicolumn{1}{c}{\textbf{CLEVR color}} \\  \cmidrule(lr){2-2}\cmidrule(lr){3-3}\cmidrule(lr){4-4}
        $p$    & \textbf{Probability mass}           & \textbf{Probability mass}           & \textbf{Probability mass}                \\\midrule
        {50}   & {94,47\%}                           & {89,69\%}                           & {94,36\%}                                \\
        {100}  & {95,16\%}                           & {92,19\%}                           & {95,33\%}                                \\
        {300}  & {93,65\%}                           & {93,61\%}                           & {94,15\%}                                \\
        {500}  & {96,57\%}                           & {92,29\%}                           & {94,15\%}                                \\
        {1000} & {95,93\%}                           & {92,93\%}                           & {93,82\%}                                \\
        \bottomrule
    \end{tabular}
    \caption{Probability masses of the \emph{attention reference resolver} for different projection sizes $p$ of the model after 20 epochs with $LSTM_e=15$ and $LSTM_o=1500$}
    \label{tab:results:attention-reference-resolver}
\end{table}

This conclusion gets even more clear, when having a look at the image, where the model identified the wrong regions.
Figure \ref{fig:visualizations_attention} shows three images for each dataset, where the model's predicted probability mass was lower than 60\%.
The errors of the models can be classified into three categories.
Most of the errors are based on non-visible target objects.
Either the target object is located outside the preprocessed cropped image, or it is partially hidden behind another object.
Unsurprisingly, the model has difficulties to focus on these regions.
However, in several cases, the models can still make use of small hints of the target object.
In the left image if figure \ref{fig:visualizations_dale-2_attention}, a corner of the target object is visible, which the model uses to 31,9\% of its attention on.
In the central image of the same figure, only the shadow of the target image is visible.
Still, the model 44,8\% of its attention onto the correct region, while the remaining probabilities are on adjacent regions.
This indicates that the model can base its decision on excluding a distractor if it doesn't match the referring expression.
This behavior is mainly visible on the 'Dale-2' dataset with only one distractor.
On the 'Dale-5' dataset, the model often focuses one on of the remaining distractors.
Hereby, the model still focuses on distractors that have attributes mentioned in the referring expressions, as the \emph{sphere} in the left image of Figure \ref{fig:visualizations_dale-5_attention} or the \emph{cyan} and \emph{cylinder} distractors in the central image.

The second category of errors includes errors in which the model confuses a distractor with the target object.
Most of the time, the model still focuses partially on target object, but the main focus lies somewhere else.
Again, in these cases, the distractor usually shares an attribute with the target object.
This is for instance shown in the right image of Figure \ref{fig:visualizations_dale-5_attention}, where the model focuses on the \emph{blue} and \emph{small cube} distractors.

Finally, errors also happen when the target region is not completely placed on the target object.
In several cases the automatic calculation of the target region around the center point doesn't include the parts of the object.
The model then doesn't actually make an error and focuses on the correct object, but attends only to adjacent regions.
These predictions are then evaluated as wrong, even though the model was attending correctly.
This can be for example seen in the left and central images of Figure \ref{fig:visualizations_colour_attention}.

\begin{figure}[ht]
    \centering
    \subfigure['Dale-2']{
        \includegraphics[width=.839\linewidth]{figures/visualization_dale-2_attention.png}
        \label{fig:visualizations_dale-2_attention}
    }
    \subfigure['Dale-5']{
        \includegraphics[width=.839\linewidth]{figures/visualization_dale-5_attention.png}
        \label{fig:visualizations_dale-5_attention}
    }
    \subfigure['CLEVR color']{
        \includegraphics[width=.839\linewidth]{figures/visualization_colour_attention.png}
        \label{fig:visualizations_colour_attention}
    }
    \caption{Examples of the \emph{attention reference resolvers'} predictions with a probability mass lower than 60\% on the 'Dale' and 'CLEVR color' datasets. The black rectangle shows the cropped section, the model is actually seeing after the image is preprocessed. The green rectangle surrounds the target region that needs to be predicted while the red regions show the actual predictions of the model. The more intense the red, the higher is the probability that the model assigned to this region.}
    \label{fig:visualizations_attention}
\end{figure}
